A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-04 analysis/paper/paper.Rmd

```{r, results='asis' }
add_stable(
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
"implementations",
"Overview of the trajectory inference methods included in this study."
)
```

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview"),
  ref_id = "evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

```{r}
add_sfig(
  figure_file(
    "robustness.svg",
    "11-evaluation_robustness"
  ),
  "robustness_comparison", 
  "The robustness of our evaluation.",
  "a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. \"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types. b) Violin plots of the overall performance on each individual dataset. For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05). c) Estimate of the size of largest subset of datasets on which the method would be the average top performer. For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether after aggregating all scores leads to the method being ranked first. This number is therefore only a        lower bound, and the actual size of largest subset could be higher."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("topology_complexity_comparison.svg", experiment_id = "8-compare_topology"),
  ref_id = "topology_complexity_comparison", 
  caption_main = "Comparing the ability of the top TI methods to detect the correct trajectory type.", 
  caption_text = "Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
)
```

@@

### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap. We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`). This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones in the network.

```{r}
add_sfig(
  figure_file('wrapping.svg', '4-method_characterisation'),
  "wrapping", 
  "A common interface for TI methods.", 
  "a) The input and output of each TI method is standardised. As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is explained in more detail.", 
  15, 
  7.5
)
```

Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`). Special conversions are denoted by an \*, and will be explained in more detail below.

@@

## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

```{r}
add_sfig(
  figure_file("score_aspects_v2.svg", experiment_id="3-score_rules/5-score_aspects"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
  "",
  15, 
  8
)
```

@@

```{r}
add_sfig(
  figure_file("metric_geodesic.svg", experiment_id="evaluation_overview"), 
  "metric_geodesic", 
  "The modified geodesic distances demonstrated on a toy example.", 
  "a) A toy example containing four milestones (W to Z) and five cells (a to e). b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model. c) The calculations made for calculating the pairwise geodesic distances. d) A heatmap representation of the pairwise geodesic distances."
)
```

@@

According to the defined common trajectory model (`r ref("fig", "evaluation_overview", "b")`), cells are also allowed to have a delayed commitment. In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network. In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network. For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$. The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.

@@

```{r}
add_sfig(
  figure_file("metric_aggregation_overview_v1.svg", experiment_id="evaluation_overview"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.",
  "",
  15, 
  5
)
```

@@

```{r}
add_sfig(
  figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview'),
  "extended_evaluation_overview",
  "Extended overview of our evaluation pipeline.", 
  "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  15, 
  7.5
)
```

>>>>

```{r, results='asis' }
add_stable(
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
"implementations",
paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
)
```

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

```{r}
add_sfig(
  fig_path = figure_file("overview_evaluation_expanded_v1.svg", "manual_figures"),
  ref_id = "overview_evaluation_expanded",
  caption_main = "Extended overview of our evaluation pipeline.", 
  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  width = 15, 
  height = 7.5
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

```{r}
add_sfig(
  figure_file(
    "robustness.svg",
    "11-evaluation_robustness"
  ),
  "robustness_comparison", 
  "The robustness of our evaluation.",
  "a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. \"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types. b) Violin plots of the overall performance on each individual dataset. For each method, the arch indicates the first method with which a method performs significantly higher (using a pairwise wilcoxon test corrected for multiple testing; p-value < 0.05). c) Estimates of the size of largest subset of datasets on which the method would be the average top performer. For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether this method ranks first when aggregating the scores over these datasets. This number is therefore only a lower bound, and the actual size of largest subset could be higher."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("topology_complexity_comparison.svg", experiment_id = "8-compare_topology"),
  ref_id = "topology_complexity_comparison", 
  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
)
```

@@

### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap. We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`). This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones in the network.

```{r}
add_sfig(
  figure_file("overview_wrapping_v1.svg", "manual_figures"),
  "overview_wrapping", 
  "A common interface for TI methods.", 
  "a) The input and output of each TI method is standardised. As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is shown in more detail.", 
  15, 
  7.5
)
```

Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "overview_wrapping", "b")`). Special conversions are denoted by an \*, and will be explained in more detail below.

@@

## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

```{r}
add_sfig(
  figure_file("aspects_v2.svg", experiment_id="manual_figures/metrics"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
  "",
  15, 
  8
)
```

@@

```{r}
add_sfig(
  figure_file("geodesic.svg", experiment_id="manual_figures/metrics"), 
  "metric_geodesic", 
  "The modified geodesic distances demonstrated on a toy example.", 
  "a) A toy example containing four milestones (W to Z) and five cells (a to e). b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model. c) The calculations made for calculating the pairwise geodesic distances. d) A heatmap representation of the pairwise geodesic distances."
)
```

@@

According to the defined common trajectory model (`r ref("fig", "ti_evaluation_overview", "b")`), cells are also allowed to have a delayed commitment. In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network. In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network. For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$. The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.

@@

```{r}
add_sfig(
  figure_file("overview_aggregation_v1.svg", experiment_id="manual_figures/metrics"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.",
  "",
  15, 
  5
)
```

@@

### 2018-04-04 analysis/paper/paper_latex.Rmd

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="evaluation_overview"),
  ref_id = "evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

```{r}
add_sfig(
  figure_file(
    "robustness.tmp.pdf",
    "11-evaluation_robustness"
  ),
  "robustness_comparison", 
  "The robustness of our evaluation.",
  "a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. \"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types. b) Violin plots of the overall performance on each individual dataset. For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05). c) Estimate of the size of largest subset of datasets on which the method would be the average top performer. For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether after aggregating all scores leads to the method being ranked first. This number is therefore only a        lower bound, and the actual size of largest subset could be higher."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("topology_complexity_comparison.tmp.pdf", experiment_id = "8-compare_topology"),
  ref_id = "topology_complexity_comparison", 
  caption_main = "Comparing the ability of the top TI methods to detect the correct trajectory type.", 
  caption_text = "Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
)
```

@@

### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap. We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`). This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones in the network.

```{r}
add_sfig(
  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
  "wrapping", 
  "A common interface for TI methods.", 
  "a) The input and output of each TI method is standardised. As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is explained in more detail.", 
  15, 
  7.5
)
```

Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`). Special conversions are denoted by an \*, and will be explained in more detail below.

@@

## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

```{r}
add_sfig(
  figure_file("score_aspects_v2.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
  "",
  15, 
  8
)
```

@@

```{r}
add_sfig(
  figure_file("metric_geodesic.tmp.pdf", experiment_id="evaluation_overview"), 
  "metric_geodesic", 
  "The modified geodesic distances demonstrated on a toy example.", 
  "a) A toy example containing four milestones (W to Z) and five cells (a to e). b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model. c) The calculations made for calculating the pairwise geodesic distances. d) A heatmap representation of the pairwise geodesic distances."
)
```

@@

```{r}
add_sfig(
  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.",
  "",
  15, 
  5
)
```

@@

```{r}
add_sfig(
  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
  "extended_evaluation_overview",
  "Extended overview of our evaluation pipeline.", 
  "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  15, 
  7.5
)
```

>>>>

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

```{r}
add_sfig(
  fig_path = figure_file("overview_evaluation_expanded_v1.tmp.pdf", "manual_figures"),
  ref_id = "overview_evaluation_expanded",
  caption_main = "Extended overview of our evaluation pipeline.", 
  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  width = 15, 
  height = 7.5
)
```

@@

```{r}
add_sfig(
  figure_file(
    "robustness.tmp.pdf",
    "11-evaluation_robustness"
  ),
  "robustness_comparison", 
  "The robustness of our evaluation.",
  "a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. \"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types. b) Violin plots of the overall performance on each individual dataset. For each method, the arch indicates the first method with which a method performs significantly higher (using a pairwise wilcoxon test corrected for multiple testing; p-value < 0.05). c) Estimates of the size of largest subset of datasets on which the method would be the average top performer. For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether this method ranks first when aggregating the scores over these datasets. This number is therefore only a lower bound, and the actual size of largest subset could be higher."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("topology_complexity_comparison.tmp.pdf", experiment_id = "8-compare_topology"),
  ref_id = "topology_complexity_comparison", 
  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
)
```

@@

### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap. We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`). This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones in the network.

```{r}
add_sfig(
  figure_file("overview_wrapping_v1.tmp.pdf", "manual_figures"),
  "overview_wrapping", 
  "A common interface for TI methods.", 
  "a) The input and output of each TI method is standardised. As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. b) The specific transformations performed by each of the wrapper functions is shown in more detail.", 
  15, 
  7.5
)
```

Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "overview_wrapping", "b")`). Special conversions are denoted by an \*, and will be explained in more detail below.

@@

## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

```{r}
add_sfig(
  figure_file("aspects_v2.tmp.pdf", experiment_id="manual_figures/metrics"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
  "",
  15, 
  8
)
```

@@

```{r}
add_sfig(
  figure_file("geodesic.tmp.pdf", experiment_id="manual_figures/metrics"), 
  "metric_geodesic", 
  "The modified geodesic distances demonstrated on a toy example.", 
  "a) A toy example containing four milestones (W to Z) and five cells (a to e). b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model. c) The calculations made for calculating the pairwise geodesic distances. d) A heatmap representation of the pairwise geodesic distances."
)
```

@@

```{r}
add_sfig(
  figure_file("overview_aggregation_v1.tmp.pdf", experiment_id="manual_figures/metrics"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.",
  "",
  15, 
  5
)
```

@@
